---
title: "Tutorial_23Sep2024"
author: "Yin Chen Wan"
date: "2024-09-21"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

# Tutorial

## Some super basic recap

### Vector

Accessing the vector content: vector[position number]:

```{r vector}
words <- c("one", "two", "three", "four", "five")
words

#Accessing each element
words[1]
words[2]
words[3]
words[4]
words[5]

#Accessing more than 1 element
words[1:3]
words[3:5]

#Putting stuff in vector (e.g. what if I want a vector with "one", "two", "three", "five"?)
#words[1:3, 5] #error #exceeds the dimension
words[1:3]
words[5]
c(words[1:3], words[5]) #try running the functions inside individually
one_two_three_five <- c(words[1:3], words[5])
one_two_three_five_four <- c(one_two_three_five, words[4])

```

### Data Frame

Since "gene_condition_source_id.txt" file is so big, I will use a small
section of it as an example:

```{r data frame}
#Read the "gene_condition_source_id.txt" file
dataFrame = read.table("gene_condition_source_id.txt", header=TRUE, sep="\t", fill=T, comment.char = "", quote = "")

#Look at the top few rows
head(dataFrame)

#save the top few rows as an example we will run
dataFrame_head <- head(dataFrame)

```

Accessing the data frame content: dataframe[row number, column number]:
*Tip is to keep checking the table itself*

```{r data frame access cell}
#Accessing each cell/element
dataFrame_head[1,1] #row 1, column 1

#going through each row of column 1
dataFrame_head[1,1]
dataFrame_head[2,1]
dataFrame_head[3,1]
dataFrame_head[4,1]
dataFrame_head[5,1]
dataFrame_head[6,1]
#now play with column 2, 3...

#going through each column of row 1
dataFrame_head[1,1]
dataFrame_head[1,2]
dataFrame_head[1,3]
dataFrame_head[1,4]
dataFrame_head[1,5]
dataFrame_head[1,6]
dataFrame_head[1,7]
dataFrame_head[1,8]
dataFrame_head[1,9]
dataFrame_head[1,10] #Oops, there are only 9 columns!
#now play with column 2, 3... 
```

Accessing data frame entire row/column:

```{r  data frame access row/column}
#Accessing row
dataFrame_head[1, ]
dataFrame_head[2, ]
dataFrame_head[3, ]

#Accessing column
dataFrame_head[, 9]
dataFrame_head[, 2]
dataFrame_head[, 5]

#Accessing column by column name 'S'
dataFrame_head$AssociatedGenes
dataFrame_head$SourceName

```

### Comparison operators

```         
<           less than
<=          less than or equal to
>           greater than
>=          greater than or equal to
==          exactly equal to
!=          not equal to
!x          not x
x | y       x OR y
x & y       x AND y e.g. if((x != 3) & (x > 0))
```

#### Comparison operators examples.

Vectors:

```{r vector comparison operators}
words <- c("one", "two", "three", "four", "five")
words == ("four")
words[words == ("four")]

numbers <- runif(30, min = 0, max = 50)
numbers > 19
numbers[numbers > 19]
```

Data frames:

```{r data frame comparison operators}
dataFrame_head$AssociatedGenes == "AAGAB"
dataFrame_head$AssociatedGenes[dataFrame_head$AssociatedGenes == "AAGAB"]
dataFrame_head$AssociatedGenes[dataFrame_head$AssociatedGenes == "TP53"]
```

Data frame example: How many diseases associated with AARS1?

```{r data frame comparson operators example}
dataFrame_head$AssociatedGenes == "AARS1"
dataFrame_head$AssociatedGenes[dataFrame_head$AssociatedGenes == "AARS1"]
dataFrame_head$DiseaseName[dataFrame_head$AssociatedGenes == "AARS1"]
```

## If else statements

if (a condition is met) { do something }

Example 1 using only "if":

```{r if else example 1}
pH <- 10

if (pH < 7) {
  print("acidic")
}

if (pH == 7) {
  print("neutral")
}

if (pH > 7) {
  print("basic")
}
```

Example 2 using "if" and "else:

```{r if else example 2}
pH <- 8

if (pH < 7) {
  print("acidic")
} else if (pH == 7) {
  print("neutral")
} else {
  print("basic")
}
```

## For loops

for (element in (a series of something)) { do something }

*You need to know how many iterations beforehand*

Super break down of a for loop using an example:

```{r for loop example 0}
words <- c("one", "two", "three", "four", "five")

i=1
words[i]

i=2
words[i]

i=3
words[i]

i=4
words[i]

i=5
words[i]

for (i in 1:length(words)) {
  print(i)
  print(words[i])
}

```

For loop example 1 (vector): Add one to every element in a vector

```{r for loop example 1}
numbers <- 100:120

#First make sure you are looping fine
for (each_number in 1:length(numbers)) {
  print(each_number)
}

for (i in 1:length(numbers)) {
  print(i)
}

for (i in 1:length(numbers)) {
  print(numbers[i])
}

#Now 'do something'
for (i in 1:length(numbers)) {
  print(numbers[i]+1)
}

for (i in 1:length(numbers)) {
  added1 <- (numbers[i]) +1
  print(added1)
}

#try something else! E.g. add 2 to every element etc.
for (i in 1:length(numbers)) {
  print(numbers[i]+2)
}

for (i in 1:length(numbers)) {
  added2 <- (numbers[i]) +2
  print(added2)
}

#Easily check your answers with vectorisation
numbers+1
```

For loop example 2 (looping through the rows of a data frame): What are
the associated diseases in the data frame?

```{r for loop example 2}
#First make sure you are looping fine
for (i in 1:nrow(dataFrame_head)) {
  print(i)
}

#Looping through the rows of the associated diseases column
for (i in 1:nrow(dataFrame_head)) {
  print(dataFrame_head[i, 5])
}

for (i in 1:nrow(dataFrame_head)) {
  print(dataFrame_head$DiseaseName[i])
}

#Easily check answer
dataFrame_head$DiseaseName
```

## While loops

counter = 1 for (a condition) { do something

counter= counter+1 }

# +1 so that it can move forward

Super break down of a loop using an example:

```{r while loop example 0}
words <- c("one", "two", "three", "four", "five")

i=1
words[i]

i=2
words[i]

i=3
words[i]

i=4
words[i]

i=5
words[i]

i=1
while (i<6) {
  print(i)
  
  i=i+1 #remove and see what happens if you don't add to the counter?
} 

i=1
while (i<6) { 
  print(i)
  print(words[i])
  
  i=i+1
}
#what if you set more than length of vector?
#vector only has 5 elements, so if it's more than 5 then it will be error
```

While loop example 1 (vector): Add one to every element in vector For
loop example 1 (vector): Add one to every element in a vector

```{r while loop example 1}
numbers <- 100:120

#First make sure you are looping fine
i=1
while (i<21) {
  print(i)
  
  i=i+1
}


i=1
while (i<21) {
  print(numbers[i])
  
  i=i+1
}

#Now 'do something'
for (i in 1:length(numbers)) {
  print(numbers[i]+1)
}
#try something else! E.g. add 2 to every element etc.

#Easily check your answers with vectorisation
numbers+1
```

While loop example 2 (how to loop through more than 1 at a time): Here
are the temperatures of 28 days (1 month/ 4 weeks). What is the average
temperature per week? \# kindly look like the assignement

```{r while loop monthly temp example}
monthly_temp <- runif(28, min=0, max = 40)
monthly_temp
```

Instead of looping through every single number, you loop through every
seven days

```{r while loop example 2}
#want to get every 7 days in 1 iteration

#first week
i=1
first_week <- monthly_temp[i: (i+7)]
first_week
length(first_week)

first_week <- monthly_temp[i: (i+6)]
length(first_week)
mean(first_week)

#2nd week
i=8
second_week <- monthly_temp[i: (i+6)]
length(second_week)
mean(second_week)

#3rd week
i=15
third_week <- monthly_temp[i: (i+6)]
length(third_week)
mean(third_week)
#what code is changing and what code is the same?

#First make sure you are looping fine
i=1
while (i<29) {
  print(i)
  
  i=i+7 #the difference between every iteration is 7 days
}


#now that the loop is working, enter what your task
i=1
while (i<29) {
  print(i)
  week <- monthly_temp[i: (i+6)]
  print(week)
  print(length(week))
  print(mean(week))
  i=i+7 #the difference between every iteration is 7 days
}
```

What if I want to put all averages into a vector? vector \<- c(vector,
things_to_put_in)

```{r appending loop output into a vector}
avg_temp <- c() #create an empty vector to put the average temperature in

#1st week
i=1
first_week <- monthly_temp[i: (i+6)]
avg_temp <- c(avg_temp, mean(first_week))
avg_temp

#2nd week
i=8
second_week <- monthly_temp[i: (i+6)]
avg_temp <- c(avg_temp, mean(second_week))
avg_temp

#3rd week
i=15
third_week <- monthly_temp[i: (i+6)]
avg_temp <- c(avg_temp, mean(third_week))
avg_temp
#what code is changing and what code is the same?

avg_temp <- c() 
i=1
while (i<29) {
  week <- monthly_temp[i: (i+6)]
  avg_temp <- c(avg_temp, mean(week))
  
  i=i+7 #the difference between every iteration is 7 days
}
avg_temp
```

## Putting if else staments into a loop

for (element in (a series of something)) { if (a condition is met) { do
something } else { do something } }

*Make sure the if else statement is working first!* if else statement in
a while loop example: Here are the temperatures of 28 days (1 month/ 4
weeks). How is the weather for each week? Assuming temperatures less
than 10 are cold, temperatures between 10-20 is warm, and temperatures
above 20 is hot. *live coding example*

```{r while loop with if/else statement}
# Example of monthly temperatures (just dummy data for 28 days)
monthly_temp <- c(15, 12, 10, 13, 17, 19, 14, 22, 23, 25, 20, 21, 18, 19, 
                  16, 14, 13, 12, 10, 9, 8, 7, 6, 5, 4, 3, 2, 1)

# 1st week
i = 1
first_week <- monthly_temp[i:(i+6)]
weather <- mean(first_week)

if (weather < 10) {
  print("cold")
} else if (weather > 20) {
  print("hot")
} else {
  print("warm")
}

# 2nd week
i = 8
second_week <- monthly_temp[i:(i+6)]  # Adjust this to cover days 8 to 14
weather <- mean(second_week)

if (weather < 10) {
  print("cold")
} else if (weather > 20) {
  print("hot")
} else {
  print("warm")
}

# 3rd week
i = 15
third_week <- monthly_temp[i:(i+6)]  # Adjust this to cover days 15 to 21
weather <- mean(third_week)

if (weather < 10) {
  print("cold")
} else if (weather > 20) {
  print("hot")
} else {
  print("warm")
}

# 4th week
i = 22
fourth_week <- monthly_temp[i:(i+6)]  # Adjust this to cover days 22 to 28
weather <- mean(fourth_week)

if (weather < 10) {
  print("cold")
} else if (weather > 20) {
  print("hot")
} else {
  print("warm")
}

#we have a partial code for this (see R chunk above)
#live coding example
```

### TIPS

**1. Write out the code for the first few iterations** Check that your
task is working. Identify what parts are the same and what parts are
changing.

**2. Check your looping is okay.**

for (i in 1:vector) { print(i) } *Make sure i is going in an order you
expect*

**3. Replace the content in loop with your if else statement**

Make sure the statement is accepting input from the loop
